---
title: "Benchmarking approaches for the analysis of single-cell expression data"
author: "Alan Murphy"
date: "Most recent update:<br> `r Sys.Date()`"
output: 
  rmarkdown::html_document: 
    theme: spacelab
    highlight: zenburn 
    code_folding: show 
    toc: true 
    toc_float: true
    smooth_scroll: true
    number_sections: false 
    self_contained: true 
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=T, message=F}
#root.dir <- here::here()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  #root.dir = root.dir
  fig.height = 6,
  fig.width = 7.00787 #178mm
)  
knitr::opts_knit$set(#root.dir = root.dir, 
                     dpi = 350)  

library(data.table)
library(cowplot)
library(wesanderson)
library(ggplot2)
library(patchwork)
```

## Background

Recently, [Zimmerman et al.](https://www.nature.com/articles/s41467-021-21038-1) 
proposed the use of mixed models over pseudobulk aggregation approaches, 
reporting improved performance on a novel simulation approach of hierarchical 
single-cell expression data. However, their reported results could not prove the
superiority of mixed models as they only calculate performance based on type 1 
error (performance of the models on non-differentially expressed genes). To 
correctly benchmark the models, a reanalysis using a balanced measure of 
performance, considering both the type 1 and type 2 errors (both the 
differentially and non-differentially expressed genes), is necessary.

This benchmark was conducted using the R/benchmark_script.R in this repository 
and an edited version of hierarchicell which returns the Matthews correlation 
coefficient performance metric as well as the type 1 error rates, uses the same 
simulated data across approaches and has checkpointing capabilities (so runs can
continue from where they left off if aborted or crashed), available at: 
https://github.com/neurogenomics/hierarchicell. 

We tested the models using the Matthews Correlation Coefficient (MCC). MCC is a 
well-known and frequently adopted metric in the machine learning field which 
offers a more [informative and reliable score on binary classification problems](https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-019-6413-7).
MCC produces scores in [-1,1] and will only assign a high score if a model 
performs well on both non-differentially and differentially expressed genes. 
Moreover, MCC scores are proportional to both the size of the differentially and
non-differentially expressed genes, so it is robust to imbalanced datasets.

The benchmark was conducted based on the average MCC from 20,000 iterations; 
50 runs for each of the 5 to 40 individuals and 50 to 500 cells at a p-value 
cut-off of 0.05 on 10,000 genes (5,000 differentially expressed genes at varying
fold changes and 5,000 non-differentially expressed genes).

## Results of benchmark analysis

We can visualise the results from the analysis, first load the data which is 
stored in this repo:

```{r}
res <- data.table::fread("./results/method_performances.csv")
print(head(res))
```

Now let's plot the results. We will split this plot to also show just the top 
four performing results (left) so the differences between these are clearer:

```{r}
#get colour palette set up - choose distinctive colours
#with similar tones for similar methods
q10 <- c(
  #orange - GEE
  "orange2",
  #yellow - modified t
  "yellow2",
  #purple - PB
  "plum2",
  "purple3",
  #gold - Tobit
  "goldenrod4",
  #green - GLMs
  "seagreen1",
  "green4",
  #blue - MAST
  "dodgerblue4",
  "cadetblue4",
  "cadetblue2"
)

all_n_ind <- c(5,10,20,30,40)
all_n_cells <- c(50,100,250,500)
tmp <- expand.grid(all_n_ind, all_n_cells)
tmp <- tmp[order(tmp$Var1),]
#create plotting column n_ind,n_cells - make factor so order maintained
res[,n_ind_n_cells:=factor(paste0(n_ind,"-\n",n_cells),
                            levels=apply(tmp, 1, paste, collapse="-\n"))]
#add descriptive method name
res[,method_des:=fcase(
      method == "MAST", "Two-part hurdle: Default",
      method == "MAST_Combat", "Two-part hurdle: Corrected",
      method == "MAST_RE","Two-part hurdle RE",
      method == "Monocle","Tobit",
      method == "ROTS","Modified t",
      method == "GLMM_tweedie","Tweedie: GLMM",
      method == "GLM_tweedie","Tweedie: GLM",
      method == "Pseudobulk_mean","Pseudobulk: Mean",
      method == "Pseudobulk_sum","Pseudobulk: Sum",
      method == "GEE1","GEE1"
  )]

mcc_res <- res[,mean(MCC),
               c("n_ind_n_cells","n_ind","n_cells","method_des")]
setorder(mcc_res,method_des)

#plot all res
plt_mcc1 <-
    ggplot(mcc_res)+
    geom_line(aes(x=n_ind_n_cells,y=V1,colour=method_des,group=method_des))+
    geom_vline(xintercept=4.5,color="grey",linetype = 3)+
    geom_vline(xintercept=8.5,color="grey",linetype = 3)+
    geom_vline(xintercept=12.5,color="grey",linetype = 3)+
    geom_vline(xintercept=16.5,color="grey",linetype = 3)+
    theme_cowplot()+
    scale_colour_manual(values=q10)+
    labs(colour='Method',y="Matthews Correlation Coefficient (MCC)",
         x="Number of Individuals - Number of cells")+
  theme(axis.text.x = element_text(size=6),
        axis.text.y = element_text(size=6),
        axis.title.x = element_text(size=9),
        axis.title.y = element_text(size=9),
        legend.text = element_text(size=9),
        legend.title = element_blank())
#plot top 4
plt_mcc2 <-
  ggplot(mcc_res[method_des %in% c("Pseudobulk: Mean","Pseudobulk: Sum",
                                    "GEE1","Tweedie: GLMM"),])+
  geom_line(aes(x=n_ind_n_cells,y=V1,colour=method_des,group=method_des),
            show.legend = FALSE)+
  geom_vline(xintercept=4.5,color="grey",linetype = 3)+
  geom_vline(xintercept=8.5,color="grey",linetype = 3)+
  geom_vline(xintercept=12.5,color="grey",linetype = 3)+
  geom_vline(xintercept=16.5,color="grey",linetype = 3)+
  theme_cowplot()+
  labs(colour='Method',y="Matthews Correlation Coefficient (MCC)",
       x="Number of Individuals - Number of cells")+
  theme(axis.text.x = element_text(size=6),
        axis.text.y = element_text(size=6),
        axis.title.x = element_text(size=9),
        axis.title.y = element_blank(),
        legend.text = element_text(size=9),
        legend.title = element_blank())+
  #get matching colours for DEG approaches from full graph
  scale_colour_manual(values=q10[c(1,3,4,7)])

plt_mcc1 + plt_mcc2 + 
  plot_layout(guides = "collect") & theme(legend.position = "bottom")
```


So it is clear that our results demonstrate that pseudobulk approaches are far 
from being "too conservative" and are, in fact, the best performing models 
based on this simulated dataset for the analysis of single-cell expression data.

## Future work

Pseudobulk approaches were also found to be the top performing approaches in a 
recent review by [Squair et al.](https://www.nature.com/articles/s41467-021-25960-2).
Notably, the pseudobulk method used here; DESeq25, performed worse than other 
pseudobulk models in Squair et al.,’s analysis and so their adoption may further
increase the performance of pseudobulk approaches on our dataset. Conversely, 
Squair et al., did not consider all models included in our analysis or the 
different forms of pseudobulk aggregation. Therefore, our results on sum and 
mean pseudobulk extend their findings and indicate that mean aggregation may be 
the best performing. 
However, you should be cognizant of the number of cells simulated by 
hierachicell, which are x10 to x100 times lower than that expected in the 
latest single-cell expression datasets hence, the results reported here may vary
on larger datasets. We speculate that the methods that noted increased 
performance as the number of cells increased (Pseudobulk: Mean and 
Pseudobulk: Sum, GEE1 and Tweedie: GLMM) would see this trend continue but this 
should be tested. Moreover, the use of simulated datasets in our analysis may 
not accurately reflect the differences between individuals that are present in 
biological datasets. Thus, despite both our results and those reported by 
Squair et al., there is still room for further analysis, benchmarking more 
models, including different combinations of pseudobulk aggregation methods and 
models, on more representative simulated datasets and biological datasets to 
identify the optimal approach. 